Project Setup

### 2024-12-10 choose_license.qmd
Rework adding data set
- Use pointblank for the data dictionary
- Create data set automatically
- Make filename of data set all lowercase

As for the reason: I think it doesn't make
sense to use datawizard for the data
dictionary, as a proper dictionary should
contain information that cannot be just
"calculated" from the data object. For
example, the information about valid values
has to come from the user, rather than from
the data set itself.

```{.bash .code-overflow-wrap filename="Terminal"}
reuse annotate --copyright="Kristen Gorman" --license="CC0-1.0" Data.csv
```

You will notice that this creates another file called `Data.csv.license` containing the relevant information:

```{.yml filename="Data.csv.license"}
SPDX-FileCopyrightText: 2024 Kristen Gorman

>>>>

```{.bash .code-overflow-wrap filename="Terminal"}
reuse annotate --copyright="Kristen Gorman" --license="CC0-1.0" data.csv
```

You will notice that this creates another file called `data.csv.license` containing the relevant information:

```{.yml filename="data.csv.license"}
SPDX-FileCopyrightText: 2024 Kristen Gorman

### 2024-12-10 make_readme.qmd
Rework adding data set
- Use pointblank for the data dictionary
- Create data set automatically
- Make filename of data set all lowercase

As for the reason: I think it doesn't make
sense to use datawizard for the data
dictionary, as a proper dictionary should
contain information that cannot be just
"calculated" from the data object. For
example, the information about valid values
has to come from the user, rather than from
the data set itself.

- `Bibliography.bib`: bibliographic references used in the manuscript
- `Data.csv`: a data set containing the simplified `palmerpenguins` data created using `write.csv(palmerpenguins::penguins, "dat.csv", row.names = FALSE)`
- `data_dictionary.html`: a dictionary to the data file, created using `create_data_dictionary.R`

>>>>

- `Bibliography.bib`: bibliographic references used in the manuscript
- `data.csv`: a data set containing the simplified `palmerpenguins` data created using `write.csv(palmerpenguins::penguins, "dat.csv", row.names = FALSE)`
- `data_dictionary.html`: a dictionary to the data file, created using `create_data_dictionary.R`

### 2024-12-10 setup.qmd
Various minor stylistic and typographic improvements
Rework adding data set
- Use pointblank for the data dictionary
- Create data set automatically
- Make filename of data set all lowercase

As for the reason: I think it doesn't make
sense to use datawizard for the data
dictionary, as a proper dictionary should
contain information that cannot be just
"calculated" from the data object. For
example, the information about valid values
has to come from the user, rather than from
the data set itself.

Then, one can open an R session by simply typing `R` into the terminal. Next, make sure that `getwd()` indicates that the working directory is `code-publishing-exercise`. If not, set it using `setwd("code-publishing-exercise")`. Then, initialize `renv`:

@@

You can now download the data set we have prepared for you and put it into your project folder: [`Data.csv`](_01_Example/Data.csv){download=""}

@@

The data set is from the package [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/)^[Created with `palmerpenguins` v0.1.1 using `write.csv(palmerpenguins::penguins, file = "Data.csv", row.names = FALSE)`.] and contains the recorded bill lengths and sex of penguins living on three islands in the Palmer Archipelago, Antarctica. It was made available under the license [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/).

@@

When publishing a data set, it is important to document the meaning (e.g., units) and possible values of its variables. This is typically done with a _data dictionary_ (also called a _codebook_). In the following, we will demonstrate how to create a simple data dictionary using the R packages [`tinylabels`](https://cran.r-project.org/package=tinylabels), [`datawizard`](https://easystats.github.io/datawizard/), and [`tinytable`](https://vincentarelbundock.github.io/tinytable/). You can install them now using:

```{.r filename="Console"}
renv::install(c(
  "tinylabels",
  "datawizard",
  "tinytable"
))
```

@@

Using `tinylabels` we can add labels to the variables of a `data.frame` in R:^[Note that the code provided does not alter the data file -- no labels will be added to `Data.csv`. The labels are only added to a (temporary) copy of the data set within R in order to create the data dictionary.]

```{r, echo = -1, eval = -2, filename="create_data_dictionary.R"}
dat <- read.csv(file.path("_01_Example", "Data.csv"))
dat <- read.csv("Data.csv")

descriptions <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)
tinylabels::variable_label(dat) <- descriptions
```

Subsequently, `datawizard` can be employed to create the data dictionary containing the name and label, but also some other information about each variable:

::: {.column-margin}
![datawizard: Easy Data Wrangling and Statistical Transformations](images/datawizard.png){width=250px}
:::

```{r, filename="create_data_dictionary.R"}
(dict <- datawizard::data_codebook(dat) |>
  subset(select = -.row_id) |>
  tinytable::tt())
```

Finally, we can store the data dictionary inside an HTML file using the R package `tinytable` and put the HTML file into the project folder as well.

::: {.column-margin}
![tinytable: Simple and Configurable Tables](images/tinytable.svg){width=250px}
:::

```{.r filename="create_data_dictionary.R"}
tinytable::save_tt(dict, output = "data_dictionary.html")
```

A human-readable data dictionary is necessary for making one's research reproducible and the example we provided demonstrates only the bare minimum. A full data documentation including measurement instruments, sampling procedures, appropriate weighting, contact information, and more information about the study can be created with the R package [`pointblank`](https://rstudio.github.io/pointblank/articles/INFO-1.html). And one could go even further by making the information machine-readable in a standardized way. We provide an optional example of that in @nte-frictionless. If you want to learn more about the sharing of research data, have a look at the tutorial "[FAIR research data management](https://lmu-osc.github.io/FAIR-Data-Management/)".

@@

This example demonstrates how the title and description of the data set, the description of the variables and their possible values are stored in a machine-readable way.

```{.r filename="Console"}
table_info <- c(
  title = "penguins data set",
  description = "Size measurements for adult foraging penguins near Palmer Station, Antarctica"
)
descriptions <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)
vals <- list(
  species = c("Adelie", "Gentoo", "Chinstrap"),
  island = c("Torgersen", "Biscoe", "Dream"),
  sex = c("male", "female"),
  year = c(2007, 2008, 2009)
)
```

@@

```{.r filename="Console"}
# Read data and create schema
dat_filename <- "Data.csv"
dat <- read.csv(dat_filename)
dat_schema <- frictionless::create_schema(dat)

@@

# Record possible values
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  if (x$name %in% names(vals)) {
    modifyList(x, list(constraints = list(enum = vals[[x$name]])))
  } else {
    x
  }
})

@@

# Create package with license info and write it
dat_package <- frictionless::create_package() |>
  frictionless::add_resource(
    resource_name = "penguins",
    data = dat_filename,
    schema = dat_schema,
    title = table_info[["title"]],
    description = table_info[["description"]],
    licenses = list(list(
      name = "CC0-1.0",
      path = "https://creativecommons.org/publicdomain/zero/1.0/",
      title = "CC0 1.0 Universal"
    ))
  )
frictionless::write_package(dat_package, directory = ".")
```

This creates the metadata file `datapackage.json` in the current directory. Make sure it is located in the same folder as `Data.csv`, as together they comprise a [data package](https://specs.frictionlessdata.io/data-package/).
:::

>>>>

Then, one can open an R session by simply typing `R` into the terminal. Next, make sure that `getwd()` indicates that the working directory is `code-publishing-exercise`. Then, initialize `renv`:

@@

```{r}
#| echo: false

write.csv(palmerpenguins::penguins, file = "data.csv", row.names = FALSE)
```

You can now download the data set we have prepared for you and put it into your project folder: [`data.csv`](data.csv){download=""}

@@

The data set is from the package [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/) (v`r packageVersion("palmerpenguins")`) and contains the recorded bill lengths and sex of penguins living on three islands in the Palmer Archipelago, Antarctica. It was made available under the license [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/).

@@

When distributing a data set, it is important to document the meaning (e.g., units) and valid values of its variables. This is typically done with a _data dictionary_ (also called a _codebook_). In the following, we will demonstrate how to create a simple data dictionary using the R package [`pointblank`](https://rstudio.github.io/pointblank/). You can install it now using:

::: {.column-margin}
![pointblank: Data Validation and Organization of Metadata for Local and Remote Tables](images/pointblank.png){width=250px}
:::

```{.r filename="Console"}
renv::install("pointblank")
```

@@

First, we write down everything we know about the data set. This includes:

- a general description of the data set
- descriptions of all columns
- valid values, where applicable

```{r, filename="create_data_dictionary.R"}
table_info <- c(
  title = "palmerpenguins::penguins",
  description = "Size measurements for adult foraging penguins near Palmer Station, Antarctica"
)
descriptions <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)

vals <- list(
  species = c("Adelie", "Gentoo", "Chinstrap"),
  island = c("Torgersen", "Biscoe", "Dream"),
  sex = c("male", "female"),
  year = c(2007, 2008, 2009)
)
```

Depending on the type of data, it may also be necessary to describe measurement instruments, sampling procedures, appropriate weighting, or contact information. In this case, as the data have already been published, we only store a reference to its source:

```{r, class.output = "r code-overflow-wrap", echo = FALSE}

# We have provided the data set as CSV file to the readers.
# Therefore, we cannot assume or require that readers have
# the R package palmerpenguins installed. Instead, we create
# the citation on our end and hide how we obtained it.

dat_source <- citation("palmerpenguins", auto = TRUE) |>
  format(bibtex = FALSE, style = "text")
dat_source |>
  gsub("\\n", " ", x = _) |>
  gsub("\"", "\\\\\"", x = _) |> # Escape any quotation marks
  paste0("dat_source <- \"", ... = _, "\"") |>
  cat()
```

Then, we use `pointblank` to create a data dictionary with this information.

```{r, filename="create_data_dictionary.R"}
vals <- sapply(vals, \(x) {
  paste0(
    "(",
    knitr::combine_words(x, and = " or ", before = "`", after = "`"),
    ")"
  )
})

dat <- read.csv("data.csv")

dict <- pointblank::create_informant(
  dat,
  tbl_name = NA,
  label = table_info[["title"]],
  lang = "en"
) |>
  pointblank::info_tabular(
    Description = table_info[["description"]],
    Source = dat_source
  ) |>
  pointblank::info_columns_from_tbl(stack(descriptions)[2:1]) |>
  pointblank::info_columns_from_tbl(stack(vals)[2:1]) |>
  pointblank::get_informant_report(
    title = "Data Dictionary for `data.csv`"
  )

dict
```

Finally, we can store the data dictionary inside an HTML file and put the HTML file into the project folder as well.

```{.r filename="create_data_dictionary.R"}
pointblank::export_report(dict, filename = "data_dictionary.html")
```

For a more elaborate introduction to `pointblank`, you can read their [Intro to Information Management](https://rstudio.github.io/pointblank/articles/INFO-1.html).

One could go even further by making the information machine-readable in a standardized way. We provide an optional example of that in @nte-frictionless. If you want to learn more about the sharing of research data, have a look at the tutorial "[FAIR research data management](https://lmu-osc.github.io/FAIR-Data-Management/)".

@@

This example demonstrates how the title and description of the data set, the description of the variables and their valid values are stored in a machine-readable way. As before, we also provide a reference to the source.

```{r, class.output = "r code-overflow-wrap", echo = FALSE}

# We have provided the data set as CSV file to the readers.
# Therefore, we cannot assume or require that readers have
# the R package palmerpenguins installed. Instead, we create
# the citation on our end and hide how we obtained it.

citation("palmerpenguins", auto = TRUE)$url |>
  paste0("dat_source <- \"", ... = _, "\"") |>
  cat()
```

@@

```{.r filename="Console"}
# Read data and create schema
dat_filename <- "data.csv"
dat <- read.csv(dat_filename)
dat_schema <- frictionless::create_schema(dat)

@@

# Record valid values
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  if (x$name %in% names(vals)) {
    modifyList(x, list(constraints = list(enum = vals[[x$name]])))
  } else {
    x
  }
})

@@

# Create package with license info and write it
dat_package <- frictionless::create_package() |>
  frictionless::add_resource(
    resource_name = "penguins",
    data = dat_filename,
    schema = dat_schema,
    title = table_info[["title"]],
    description = table_info[["description"]],
    licenses = list(list(
      name = "CC0-1.0",
      path = "https://creativecommons.org/publicdomain/zero/1.0/",
      title = "CC0 1.0 Universal"
    )),
    sources = list(list(
      title = "CRAN",
      path = dat_source
    ))
  )
frictionless::write_package(dat_package, directory = ".")
```

This creates the metadata file `datapackage.json` in the current directory. Make sure it is located in the same folder as `data.csv`, as together they comprise a [data package](https://specs.frictionlessdata.io/data-package/).
:::
